Trap spaces of Boolean networks are conflict-free siphons of their Petri net encoding

### 2023-02-18 manuscripts/TCS2022/tcs2022.tex
position with ensembles of Boolean networks

Second, we consider fixed points in Boolean networks.
Let \(s\) be a fixed point of a Boolean network \(\mathcal{N}\).
We have a subspace \(m\) corresponding to \(s\) as follows: \(\forall v \in V, m(v) = s(v)\), i.e., all nodes are fixed in \(m\).
Clearly, \(s\) is a trap set of \(\mathcal{N}\) regardless of the update scheme.
Hence, \(m\) is a trap space of \(\mathcal{N}\).
In addition, since \(\left\vert S_{\mathcal{N}}[m]\right\vert = 1\), \(m\) is also a minimal trap space.
To compute all fixed points of \(\mathcal{N}\), we can add more constraints to the encoded ASP characterizing all conflict-free siphons (equivalently trap spaces).
For every \(v \in V\), we add to the encoded ASP the rule
\[
  \verb|p-v ; n-v.|
\]
that ensures that for every conflict-free siphon \(S\), it contains either \(\verb|p-v|\) or \(\verb|n-v|\) for every \(v \in V\).
Equivalently, the trap space corresponding to \(S\) is always a fixed point.
Now, the set of answer sets of the encoded ASP is equivalent to the set of fixed points of \(\mathcal{N}\).
In particular, when solving the encoded ASP using an ASP solver, we do not need to use the built-in option for computing set-inclusion maximal answer sets.
Note that we can also build another ASP characterizing all fixed points of \(\mathcal{N}\) based on the equivalence between a fixed point of \(\mathcal{N}\) and a deadlock of its Petri net encoding~\cite{chatain2014characterization}.
This approach may give a more compact ASP\@.

@@

Non-deterministic behavior is problematic for Boolean networks with the synchronous update scheme, which are unable to represent the possibility of more than one next state.
It can be addressed by using Boolean networks with asynchronous update schemes because of their non-determinism.
However, multiple update functions may be considered for each node, which is problematic for existing analysis methods in Boolean networks.
For example, if the truth table of a node have \(k\) unknown-output rows, then this node may have \(2^k\) possible Boolean functions and the analysis methods such as \texttt{PyBoolNet} and \texttt{mpbn} need to consider \(2^k\) possible Boolean networks due to their inputs are Boolean networks with complete Boolean functions.
Moreover, one interesting question maybe raised from systems biologists is if there is a common behavioral property (e.g., minimal trap space, attractor) among all possible combinations.
To answer this question, the existing analysis methods maybe need to consider \(2^k\) possible Boolean networks as well.

@@

\begin{example}
Consider an incomplete Boolean network \(\mathcal{N} = (V, F)\) (mentioned in~\cite{Steggles2006}), where \(V = (g_1, g_2, g_3)\) and \(F = (f_1, f_2, f_3)\) with \(f_1 = g_2, f_3 = \neg g_1\). Herein, \(f_2\) is an incomplete function and the partial truth table is as follows.
\begin{center}
	\begin{tabular}{rr|r}
		\toprule
		$g_1$ & $g_3$ & $g_2$ \\ \midrule
		0 & 0 & 0 \\
		0 & 1 & 0 \\
		1 & 0 & ? \\
		1 & 1 & 1 \\
		\bottomrule
	\end{tabular}
\end{center}
We have two possible update functions for \(g_2\): \(f_2 = g_1 \land g_3\) for the case that \(? = 0\) and \(f_2 = g_1\) for the case that \(? = 1\).
Accordingly, we have two possible Boolean networks.
The first one has one minimal trap space: 001.
The second one has two minimal trap spaces: 001 and 110.
Both have a common minimal trap space: 001.
The asynchronous dynamics of the incomplete Boolean network is shown in Figure~\ref{fig:incomplete-dynamics-new-PN}a.
The result Petri net following our idea is given in Figure~\ref{fig:incomplete-dynamics-new-PN}b.
Herein, \(t^3_{g_2}\) and \(t^4_{g_2}\) are the two conflicting transitions representing the unknown-output row.
This Petri net has only one maximal conflict-free siphon \(\{p_{g_1}, p_{g_2}, \overline{p}_{g_3}\}\) corresponding to the common minimal trap space 001.

@@

In summary, we can still perform meaningful analysis on the resulting Petri net.
Note that the new Petri net may have more transitions, however the number of additional transitions is only \(O(2k)\) where \(k\) is the number of unknown-output rows in the truth tables.
This is actually a significant advantage, since we do not need to consider \(2^k\) possible Boolean networks.
It is worth noting that the approach proposed in~\cite{Steggles2006} is only shown to deal with the case that \(k = 1\), thus it is not clear how this approach can handle the case that \(k > 1\).
Furthermore, as more data becomes available for the underlying biological system, the Petri net model can be refined to reduce the
amount of non-determinism (i.e., the number of conflicting transitions) it contains.
Hence, Petri nets provide an interesting means of documenting the development of knowledge for a biological system.

>>>>

Second, we consider fixed points in Boolean networks.
Let \(s\) be a fixed point of a Boolean network \(\mathcal{N}\).
We have a subspace \(m\) corresponding to \(s\) as follows: \(\forall v \in V, m(v) = s(v)\), i.e., all nodes are fixed in \(m\).
Clearly, \(s\) is a trap set of \(\mathcal{N}\) regardless of the update scheme.
Hence, \(m\) is a trap space of \(\mathcal{N}\).
In addition, since \(\left\vert S_{\mathcal{N}}[m]\right\vert = 1\), \(m\) is also a minimal trap space.
To compute all fixed points of \(\mathcal{N}\), we can add more constraints to the encoded ASP characterizing all conflict-free siphons (equivalently trap spaces).
For every \(v \in V\), we add to the encoded ASP the rule
\[
  \verb|p-v; n-v.|
\]
that ensures that for every conflict-free siphon \(S\), it contains either \(\verb|p-v|\) or \(\verb|n-v|\) for every \(v \in V\).
Equivalently, the trap space corresponding to \(S\) is always a fixed point.
Now, the set of answer sets of the encoded ASP is equivalent to the set of fixed points of \(\mathcal{N}\).
In particular, when solving the encoded ASP using an ASP solver, we do not need to use the built-in option for computing set-inclusion maximal answer sets.
Note that we can also build another ASP characterizing all fixed points of \(\mathcal{N}\) based on the equivalence between a fixed point of \(\mathcal{N}\) and a deadlock of its Petri net encoding~\cite{chatain2014characterization}.
This approach may give a more compact ASP\@.

@@

Non-deterministic behavior is problematic for Boolean networks with the synchronous update scheme, which are unable to represent the possibility of more than one next state.
It can be addressed by using Boolean networks with asynchronous update schemes because of their non-determinism.
However, multiple update functions may be considered for each node, which is problematic for existing analysis methods in Boolean networks.
For example, if the truth table of a node have \(k\) unknown-output rows, then this node may have \(2^k\) possible Boolean functions and the analysis methods such as \texttt{PyBoolNet} and \texttt{mpbn} need to consider \(2^k\) possible Boolean networks due to they require complete Boolean functions for their underlying processing.
Moreover, one interesting question maybe raised from systems biologists is if there is a common behavioral property (e.g., minimal trap space, fixed point, attractor) among all possible combinations.
To answer this question, the existing analysis methods maybe need to consider \(2^k\) possible Boolean networks as well.

@@

\begin{example}
Consider an incomplete Boolean network \(\mathcal{N} = (V, F)\) (mentioned in~\cite{Steggles2006}), where \(V = (g_1, g_2, g_3)\) and \(F = (f_1, f_2, f_3)\) with \(f_1 = g_2, f_3 = \neg g_1\). Herein, \(f_2\) is an incomplete function and the partial truth table is as follows.
\begin{center}
	\begin{tabular}{rr|r}
		\toprule
		$g_1$ & $g_3$ & $g_2$ \\ \midrule
		0 & 0 & 0 \\
		0 & 1 & 0 \\
		1 & 0 & ? \\
		1 & 1 & 1 \\
		\bottomrule
	\end{tabular}
\end{center}
We have two possible update functions for \(g_2\): \(f_2 = g_1 \land g_3\) for the case that \(? = 0\) and \(f_2 = g_1\) for the case that \(? = 1\).
Accordingly, we have two possible Boolean networks.
The first one has one minimal trap space: 001.
The second one has two minimal trap spaces: 001 and 110.
Both have a common minimal trap space: 001.
The asynchronous dynamics of the incomplete Boolean network is shown in Figure~\ref{fig:incomplete-dynamics-new-PN}a.
The resulting Petri net following our idea is given in Figure~\ref{fig:incomplete-dynamics-new-PN}b.
Herein, \(t^3_{g_2}\) and \(t^4_{g_2}\) are the two conflicting transitions representing the unknown-output row.
This Petri net has only one maximal conflict-free siphon \(\{p_{g_1}, p_{g_2}, \overline{p}_{g_3}\}\) corresponding to the common minimal trap space 001.

@@

In summary, we can still perform meaningful analysis on the resulting Petri net.
Note that the new Petri net may have more transitions, however the number of additional transitions is only \(O(2k)\) where \(k\) is the number of unknown-output rows in the truth tables.
This is actually a significant advantage, since we do not need to consider \(2^k\) possible Boolean networks.
It is worth noting that the approach proposed in~\cite{Steggles2006} is only shown to deal with the case that \(k = 1\), thus it is not clear how this approach can handle the case that \(k > 1\).
Moreover, in the context of Boolean models construction of biological systems from prior knowledge and experimental data, the work by~\cite{chevalier2020synthesis} also deals with some kind of uncertainty.
It uses ASP to represent all possible Boolean networks possessing the same given property, where each solution of this ASP is equivalent to a distinct Boolean network.
Then it samples through that solution space with heuristics to select \emph{diverse} Boolean networks, and proceed further simulations or analysis on every individual Boolean network.
Compared to our Petri net approach presented above, it still requires to consider multiple Boolean networks as well as it is only applicable for locally-monotonic Boolean networks~\cite{chevalier2020synthesis}.

To this end, as more data becomes available for the underlying biological system, the Petri net model can be refined to reduce the
amount of non-determinism (i.e., the number of conflicting transitions) it contains.
Specifically, if we know the output for an unknown-output row of a node, we can remove one of the two conflicting transitions corresponding to this unknown-output row.
Hence, Petri nets provide an interesting means of documenting the development of knowledge for a biological system, and our Petri net-based approach can be more useful in the model refinement loop.
